### 2011-06-07 thesis/novelty.tex
Changed TikZ styles to draw factor graphs.
Several changes on novelty detection on the conceptual map. Still WIP.

Assume now that everything the conceptual layer knows holds true except the
class of a single variable $a$, which the system possess no information on. 
In that case, the correct approach is to replace the factors associated with
$a$ with factors representing the prior-information the system has on
the unconditional distribution of those.
Let this new graph with unconditional factors on the variable
$a$ be $U_a$. $P(x)$ is then approximated by $P_{U_a}(x)$.

Note that in case no prior-information is available on the unconditional
factors this means replacing them by a uniform factor\footnote{In case the
unconditional factors are uniform, this is the same as omitting the factors
from the graph structure.}. In case some information is known
(i.e. unlabelled data) the correct approach is to incorporate it into the
unconditional factors using the principle of maximum entropy as seen on
\autoref{sec:unlabelled-data}.

\begin{figure}[h]
\centering
\subfloat[Factor graph modelling the probability distribution in the case both
          the variables types and graph structure hold true.]{
\begin{tikzpicture}
\node [matrix,matrix anchor=mid, column sep=20pt, row sep=10pt,ampersand replacement=\&] {
\&
\&
\node (room3) [latent] {$R_3$}; \& \\
\& \& \& \\
\&
\node (room1) [latent] {$R_1$}; \& \&
\node (room2) [latent] {$R_2$}; \\
\& \& \& \\
\node (shape1f) [factor] {}; \&
\node (appearance1f) [factor] {}; \&
\node (object1f) [factor] {}; \&
\node (factor2) [factor] {}; \\
\node (shape1l) [obs] {$S_p$}; \&
\node (appearance1l) [obs] {$A_p$}; \&
\node (object1l) [obs] {$O_p$}; \&
\node (prop2) [obs] {$X_p$}; \\
};

@@

\plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
\plate{places2}{(factor2)(prop2)}{$\dots$}{};

\end{tikzpicture} %
          }
\qquad\qquad
\subfloat[Factor graph modelling the probability distribution in the case
          where only variable $a$ does not conforms to any class the system
          has learned before.]{
\begin{tikzpicture}
\node [matrix,matrix anchor=mid, column sep=20pt, row sep=10pt,ampersand replacement=\&] {
\&
\&
\node (room3) [latent] {$R_3$}; \& \\
\& \& \& \\
\&
\node (room1) [latent] {$R^N_1$}; \& \&
\node (room2) [latent] {$R_2$}; \\
\& \& \& \\
\nofactor {shape1f}{}{$U$} {right=0pt}; \&
\nofactor {appearance1f}{} {$U$} {right=0pt}; \&
\nofactor {object1f}{} {$U$} {right=0pt}; \&
\node (factor2) [factor] {}; \\
\node (shape1l) [obs] {$S_p$}; \&
\node (appearance1l) [obs] {$A_p$}; \&
\node (object1l) [obs] {$O_p$}; \&
\node (prop2) [obs] {$X_p$}; \\
};

\draw [-] (room1) -- (room2) node (r12f) [midway,factor] {};
\draw [-] (room1) -- (room3) node (r13f) [midway,factor] {};
\draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};
\node (captr12f) [above=0pt of r12f] {\footnotesize{$U$}};
\node (captr13f) [above=0pt of r13f] {\footnotesize{$U$}};

\plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
\plate{places2}{(factor2)(prop2)}{$\dots$}{};

\end{tikzpicture} %
}

\caption{\label{fig:nd-single-variable}Graphs used to model novelty detection on
         a class of a single variable $R_1$, assuming all the other variables
         classes and graph structure are correctly modelled by the agent
         knowledge.}
\end{figure}

Having now defined $P_G(x)$ and $P_{U_a}$ as models models for
$P(x|\overline{novel})$ and $P(x)$ (\autoref{fig:nd-single-variable}),
$P_G(x)/P_{U_a}(x)$ is used to define an order-relation on the novelty of $a$.
Now all that is missing is a threshold $t_A$ to trigger on.
Such a threshold can be obtained by calculating which value of $t_A$
triggers a desired amount of the labelled data.
Also, since there is no reason to distinguish between different instances of
a variable type, only one threshold by variable type needs to be trained.

@@

where $t_A$ is a threshold for variables of type $A$, $P_G(x)$ denotes the
probability of the sensed variables $x$ being generated by a distribution
where $a$ is known, and $P_{Ua}$ denotes the probability of $x$ being generated
by a distribution where only $a$ is novel.

>>>>

\begin{figure}[h]
\centering
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {$c$}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent] {$a$}; \& \&
    \node (room2) [latent] {$b$}; \\
    \& \& \& \\
    \node (shape1f) [factor] {}; \&
    \node (appearance1f) [factor] {}; \&
    \node (object1f) [factor] {}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

@@

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}

\caption{Factor graph modelling the probability distribution in the case both
          the variables types and graph structure hold true.}
\end{figure}

Assume now that everything the conceptual layer knows holds true except the
class of a single variable $a$, which the system has no information on. 
In that case, the correct approach is to replace the factors associated with
$a$ with a single factor connecting all the variables dependent on it as seen
on Figure~\ref{fig:novel-a-correct}.

Such an highly connected factor is hard to model. Though assuming the structure
holds true, it is possible to approximate that highly connected factor with a
variable $a^*$ with $K$ states.\footnote{Ideally those $K$ states would be all
the classes of $a$.}
Assuming no information is known on the unconditional distribution, the factors
dependent on $a^*$ should be modelled with uniform factors and $K$ can be set to
$1$.

Let $U_a$ be this new graph based on $G$ with variable $a$ replaced by $a^*$.
$P_{U_a}(x)$ models then the unconditional distribution of the feature set $x$
without the assumption that $a$ is one of the known classes and so is used as
an approximation for $P(x)$.

\begin{figure}[h]
\centering
\subfloat[\label{fig:novel-a-correct}Graphical model able to correctly represent
          the unconditional probability when variable $a$ can be novel.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \& \& \node (room3) [latent] {$c$}; \& \\
    \& \& \& \\
    \&
    \node (room1) [factor] {}; \& \&
    \node (room2) [latent] {$b$}; \\
    \& \& \& \\
    \&
    \&
    \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };

  \draw [-] (room1) -- (shape1l);
  \draw [-] (room1) -- (appearance1l);
  \draw [-] (room1) -- (object1l);
  \draw [-] (room2) -- (prop2);

  \draw [-] (room1) -- (room2);
  \draw [-] (room1) -- (room3);
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}
\qquad
\subfloat[By using independent features for modelling the unconditional
          distribution an highly simplified graph is obtained.]{
\begin{tikzpicture}
  \node [matrix,matrix anchor=mid, column sep=15pt, row sep=10pt,ampersand replacement=\&] {
    \&
    \&
    \node (room3) [latent] {$c$}; \& \\
    \& \& \& \\
    \&
    \node (room1) [latent,dashed] {$a^*$}; \& \&
    \node (room2) [latent] {$b$}; \\
    \& \& \& \\
    \nofactor {shape1f}{}{$U$} {right=0pt}; \&
    \nofactor {appearance1f}{} {$U$} {right=0pt}; \&
    \nofactor {object1f}{} {$U$} {right=0pt}; \&
    \node (factor2) [factor] {}; \\
    \node (shape1l) [obs] {$S_p$}; \&
    \node (appearance1l) [obs] {$A_p$}; \&
    \node (object1l) [obs] {$O_p$}; \&
    \node (prop2) [obs] {$X_p$}; \\
  };
  \draw [-] (shape1f) -- (shape1l);
  \draw [-] (appearance1f) -- (appearance1l);
  \draw [-] (object1f) -- (object1l);
  \draw [-] (room2) -- (factor2) -- (prop2);
  \draw [-,draw=none] (room1) -- (room2) node (r12f) [midway,factor] {};
  \draw [-,draw=none] (room1) -- (room3) node (r13f) [midway,factor] {};
  \draw [-] (room2) -- (room3) node (r23f) [midway,factor] {};
  \draw [-] (room2) -- (r12f);
  \draw [-] (room3) -- (r13f);
  \node (captr12f) [above=0pt of r12f] {\footnotesize{$U$}};
  \node (captr13f) [above=0pt of r13f] {\footnotesize{$U$}};

  \draw [-,dashed] (room1) -- (shape1f);
  \draw [-,dashed] (room1) -- (appearance1f);
  \draw [-,dashed] (room1) -- (object1f);
  \draw [-,dashed] (room1) -- (r12f);
  \draw [-,dashed] (room1) -- (r13f);

  \begin{pgfonlayer}{background}
    \plate{places1}{(shape1f)(shape1l)(object1l)}{$\forall p \in places(room1)$}{};
    \plate{places2}{(factor2)(prop2)}{$\dots$}{};
  \end{pgfonlayer}
\end{tikzpicture}
}
\caption{Factor graph modelling the probability distribution in the case
         where only variable $a$ is not necessarily a class the system has
         learned before.}
\end{figure}

Having now defined $P_G$ and $P_{U_a}$ as models models for
$P(x|\overline{novel})$ and $P(x)$, the ratio $P_G(x)/P_{U_a}(x)$ is used to
define an order-relation for the novelty of $a$.
Now all that is missing is a threshold $t_A$ to trigger on.
Such a threshold can be obtained by calculating which value of $t_A$
triggers a desired amount of the labelled data.
Also, since there is no reason to distinguish between different instances of
a variable type, only one threshold per variable type needs to be trained.

@@

where $t_A$ is a threshold for variables of type $A$, $P_G(x)$ denotes the
probability of the sensed variables $x$ being generated by a distribution
where $a$ is known, and $P_{Ua}$ denotes the probability of $x$ being generated
by a distribution where only $a$ is not necessarily a class the system has
learned before.

\subsection{Novelty Detection as a MAP Operation}
Having interest in integrating the presented method to detect novelty of a
variable $a$ on a single graphical model some more steps can be taken.

Let the known states of $a$ be $\{A_1,\dotsc, A_{n}\}$ and the unconditional variable
$a^*$ be modelled with states $\{A^*_1,\dotsc, A^*_{k}\}$. I.e. if $k=1$ then the
unconditional variable is being modelled with an assumption on features that do
not dependent on the class.

First, extend variable $a$ with new states $A_{n+k}$

By incorporating novelty information back into the graphical model the presented
method allows
